A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-02-08 paper/paper.Rmd
placement of tables
width of printed data on page 5 fixed

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, `new_data_qnames`, does not comply with these rules as it is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year. The raw data consequently has a large number of columns (716 to be specific). The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data output is shown below.

@@

Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse]:`tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

>>>>

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, `new_data_qnames`, does not comply with these rules as it is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year. The raw data consequently has a large number of columns (716 to be specific). The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data shows:

@@

Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

### 2022-02-08 paper/paper.tex
placement of tables
width of printed data on page 5 fixed

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (716 to be specific). The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data output is shown below.

\begin{verbatim}
#> 'data.frame':    12686 obs. of  716 variables:
#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 9 10 ...
#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA 214 NA ...
#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA NA NA ...
#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA NA NA ...
#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA NA NA ...
#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA NA NA ...
#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 200 318 ...
#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA NA NA ...
#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA NA NA ...
#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA NA NA ...
#>   [list output truncated]
\end{verbatim}

Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019):\texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{CENSORED/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
Male & 1,000 (15.62\%) & 1,613 (25.19\%) & 3,790 (59.19\%) & 6,403 (100.00\%)\\
Female & 1,002 (15.95\%) & 1,561 (24.84\%) & 3,720 (59.21\%) & 6,283 (100.00\%)\\
\midrule
Total & 2,002 (15.78\%) & 3,174 (25.02\%) & 7,510 (59.20\%) & 12,686 (100.00\%)\\
\bottomrule
\end{tabular}
\end{table}

@@

>>>>

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (716 to be specific). The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data shows:

\begin{verbatim}
#> 'data.frame':    12686 obs. of  716 variables:
#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA ...
#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA ...
#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA ...
#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA ...
#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA ...
#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 ...
#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA ...
#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA ...
#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA ...
#>   [list output truncated]
\end{verbatim}

Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{CENSORED/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

@@

\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
Male & 1,000 (16\%) & 1,613 (25\%) & 3,790  (59\%) & 6,403 (100\%)\\
Female & 1,002 (16\%) & 1,561 (25\%) & 3,720  (59\%) & 6,283 (100\%)\\
\midrule
Total & 2,002 (32\%) & 3,174 (50\%) & 7,510 (118\%) & 12,686 (200\%)\\
\bottomrule
\end{tabular}
\end{table}
